# multiple regression models



```{r eval=TRUE, include=TRUE}
usbeer <- read.csv("/cloud/project/data/beernew.txt")
attach(usbeer)
head(usbeer)

```


### additivie decomposition model with a fourth-degree polynomial trend

Seasonal indices add to 0 for all t. 
$$
\begin{equation}
  y_t = \beta_0 + \beta_2t^2 +  \beta_3t^3 + \beta_4t^4 + S_t + \epsilon_t
\end{equation}
$$

R creates dummies. 

```{r}
fmonth<-as.factor(month)
levels(fmonth)
class(time)
time <- as.numeric(time)
class(time)
model1<- lm(beer~time+I(time^2)+I(time^3)+I(time^4)+fmonth);summary(model1)
```

examine residuals

```{r}
plot(ts(resid(model1),start=c(1987,1),freq=12),xlab="time",ylab="residu als",main="Residuals of Model 1")
```

normal quantile plot of residuals. tails are a bit long relative to normaility. But not much of a problem. Null of Shapiro Wilk is normality of residuals

```{r}
qqnorm(resid(model1))
qqline(resid(model1))
shapiro.test(resid(model1))
```

Calculate and interpret estimated seasonal indices from the model. Selects estimated intercpet coefficients and the estimated coefficients for the fmonth dummies, then calculates the seasonal index estimates from these. In June, production is 2.072 million barrels above the level of the trends... 

```{r}
b1<-coef(model1)[1]
b2<-coef(model1)[6:16]+b1
b3<-c(b1,b2)
seas<-b3-mean(b3)
seas

```


Variable seas to time series clas sand plot indices 

```{r}
seas.ts<-ts(seas)
plot(seas.ts,ylab="seasonal indices",xlab="month")
```


Partial F tests and hypothesis. Partial F test tests the null that some of the regression coefficients are simultaneously 0. Fir two models, one with all regressers, and a reduces. In the latter, omit variables corresponding to the parameters which are hypothesized to be 0.

Suppose T is number of observations. The F stat for the partial F test can be calculated from the R squared of the two models. The F statistic is:
$$
\begin{equation}
 F_{r,T-k-1} = \frac{((R^2(full)-R^2(reduced))/r}{(1-R^2(full))/(T-k-1)}
\end{equation}
$$

where numerator DF = r and T-k-1 denominator degrees of freedom. Reject null if partial F is large (p value small). 

```{r}

model2<-lm(beer~time+I(time^2)+I(time^3)+I(time^4))
summary(model2)
summary(model1)$r.squared
summary(model2)$r.squared
```
$$
\begin{equation}
 F_{11,356} = \frac{((0.92366-0.04045)/11}{(1-0.92366)/(356)} = 374.43
\end{equation}
$$



confirm with test

```{r}
anova(model2,model1)
```

Do residuals suggest that the model disturbance term forms an uncorrelated sequence? Measure lag 1 correlation of the residuals, correlation between residuals which are directly adjacent in time. Lag one correlation not significant.

```{r}
r<-resid(model1)[2:372]
r1<-resid(model1)[1:371]
cor.test(r,r1)
```

Autocorrelation structure of residuals by estimates of the correlations at multiple lags. See significant residual correlations at many lags (significance is indicates by extention beyond the blue dashed lines). 

```{r}
acf(ts(resid(model1)),36)
```

### Additive decomposition model with cosine and sine seasonal dummies and inclusion of an intercept

$$
\begin{equation}
  y_t = \beta_0 + \beta_1t +  \beta_2t^2 + \beta_3t^3 + \beta_4t^4 + \sum_{j=1}^{5}(\gamma_j cos\frac{2\pi jt}{12} + \delta_j \frac{2\pi jt}{12}) + \gamma_6(-1)^{t} + \epsilon_t
\end{equation}
$$

Seasonal is now written as the sum of a fundamental component with period 12 and its five overtones. The jth component has amplitude

$$
\sqrt{\gamma_j^2 + \delta_j^2}, j = 1,...,5
$$


```{r}
cosm<-matrix(nrow=length(time),ncol=6)
sinm<-matrix(nrow=length(time),ncol=5)
for(i in 1:5){cosm[,i]<-cos(2*pi*i*time/12)
 sinm[,i]<-sin(2*pi*i*time/12)}
cosm[,6]<-cos(pi*time)

model6<- lm(beer~time+I(time^2)+I(time^3)+I(time^4)+cosm[,1]+sinm[,1]+cosm[,2]+sinm[,2]+cosm[,3]+sinm[,3]+cosm[,4]+sinm[,4]+cosm[,5]+sinm[,5]+cosm[,6])
options(digits=10)
summary(model6)
```





### Australian beer

```{r}
ausbeer <- read.csv("/cloud/project/data/beeraustralia.txt")
attach(ausbeer)
head(ausbeer)
fmonth<- as.factor(month)
plot(ts(beer,start=c(1956,1),freq=12),xlab="time",ylab="beer production",main="Australia monthly beer production")
time<-as.numeric(1:length(beer)) 
modela1<-lm(log(beer)~poly(time,5)+fmonth);summary(modela1) 

b1<-coef(modela1)[1]
b2<-coef(modela1)[7:17]+b1
b3<-c(b1,b2)
seasa1<-exp(b3-mean(b3))
seasa1
usseas.ts<-ts(exp(seasa1)) 
ausseas.ts<-ts(seasa1) 
ausseas.ts<-c(ausseas.ts,ausseas.ts)[7:18] 
plot(usseas.ts,ylim=c(0.8,1.3),xlab="month",ylab="seasonal 
indices",main="comparison of estimated seasonal 
indices",col="green",lwd=2) 
lines(ausseas.ts,col="red",lwd=2) 
legend("topleft",legend=c("US","Australia"),col=c("green","red"),lty=1,
cex=0.8) 
acf(ts(resid(modela1)))



modela2<-lm(dlogbeer~fmonth);summary(modela2) 
plot(modela2, which = 2)
dlogbeer[316:319]
beer[316:319]
modela3<-lm(dlogbeer~fmonth+obs318);summary(modela3)
plot(modela3, which = 2)

b1<-coef(modela3)[1] 
b2<-coef(modela3)[2:12]+b1 
b3<-c(b1,b2) 
x<-b3-mean(b3) 
```

Getting the changes in the log S values. 
```{r}
s12<-0 
for(j in 2:12){ 
 xsub<-x[j:12] 
 s12<-s12+sum(xsub) 
 } 
s12<-s12/12 
 s<-c(rep(0,times=12)) 
 s[12]<-s12 
 for(j in 1:11){ 
 xsub<-x[1:j] 
 s[j]<-s[12]+sum(xsub) 
 } 
s<-exp(s) 
s 
 
```

```{r}

modela4<-lm(log(beer)~poly(time,5)+fmonth+obs317);summary(modela4) 

b1<-coef(modela4)[1] 
b2<-coef(modela4)[7:17]+b1
b3<-c(b1,b2) 
seasa4<-exp(b3-mean(b3))
seasa4

## May has the biggest difference
options(digits=4) 
cbind(seasa1,s,seasa4) 



```

Two cosine sin pairs with frequencies 0.348 and 0.432. Polynomial of degree 5, the fmonth that is defined, the dummy for the outlier, and the two cosine sin pairs. Cosine sin pairs are significant. The 348 pair is highly sig. 432 pair also sig.

In autocorrelation plots, there is additional seasonal structure in the beer data, and that is dynamic structure. We have extracted the static, and that is dynamic. That lag 1 is suggesting that there is a tiny bit of trend that we haven't captured with the model. There is a little bit of dynamic seasonal structure. 

```{r}
c348<-cos(0.696*pi*time);s348<-sin(0.696*pi*time) 
c432<-cos(0.864*pi*time);s432<-sin(0.864*pi*time) 
 
modela5<-
lm(log(beer)~poly(time,5)+fmonth+obs317+c348+s348+c432+s432);summary(modela5)

acf(ts(resid(modela5)))

```





