# distributed lag models

Let y be a dependent variable and x an independent variable. The model

$$
y = \alpha +\sum_{k=0}^{\infty}\beta_kx_{t-k} +\epsilon_t
$$

is called a distributed lag model. The effect of the IV on the DV is distributed over a number of lags (e.g., y is sales and x is advertising). 

Weighted sum of the value of the X's at time T and at previous times. Number of terms in the sum is infinite. Current and past advertising influences sales today. The x's are being lagged. X_t - 1,2, etc. When k=0, then we have X_t.

A special case (geometric decay) is given by

$$
\beta_k = \beta\gamma^k
$$

Some constnat Beta, and a gamma raised to the power k. If gamam is 1/2, then the weights are beta when k=0, then a half beta, quarter beta, 1/8 beta, etc. If the weights decrease in this geometric pattern, we call this geometric decay. 

For this case the model becomes 

$$
y_t = \alpha + \beta(x_t + \gamma*X_{t-1}
$$

The backward shift operator B is defined by...
$$
$$

If you operate with B3, you shift backwards 3 times... Let's substitue the backwards shift operator into the expression. Then, for geometric decay, the model is ...

$$
y_t = \alpha +\beta(1+\gamma\beta + \gamma^2\beta^2+...) \\
=\alpha+\beta(1-\gamma\beta)^{-1}x_t + \epsilon_t
$$

by use of the geometric series summation. Now multiply on both sides by the oeprator 1-$\gamma\beta$. This yields

$$
y_t = (1-\gamma)\alpha + \gamma{y_{t-1}} + \beta{x_t} + \epsilon_t - \gamma\epsilon_{t-1}
$$
or

$$
y_t = \alpha' + \gamma{y_{t-1}} + \beta{x_t} + \epsilon'_t
$$

when geometric decay prevails to define the Beta_k's. When the betas decrease in a geometric fashion, then you can re-write the model in this way. Second term is the auto-regressive term. Only have contemporaenous x. Also has an altered error structure. The original error structure is just epsilon_t.

The lag one value of the DV appears on the right hand side as an explanatory variables, and the x variable appears only contemperaneously. The structure is first order auto-regressive structure. It is an arx model. Autoregressive becayse y is regressed back on itself, and the x denote the fact that there is an explanatory x variable. The epsilon is different from the original epsilon.

When geometric decay prevails, there is this simpler way to write the model. 

Notes: the x's have different lags. There is a lot of correlation. High degree of collinearity makes it difficult to interpret the beta coefficient. Consider a simpler formulation if geometric decay prevails.  


Analysis: how to modify the data so that we can get reduction to white noise when we fit a model.


```{r}
lydia <- read.csv("/cloud/project/data/Lydia.txt")
attach(lydia)
head(lydia)
sales<- lydia$sales
ad<- lydia$advrtsng
plot(ts(sales,start=1907),ylim=c(0,4000),
     xlab="year",main="Sales and
Advertising vs. Year",lty=1,lwd=2,col="red")
lines(ts(ad,start=c(1907,1)),lty=2,lwd=2,col="blue")

```

Sales and advertising series have similar traces. Advertising series appears to lag the sales series by perhaps a year in its pattern.

The model below is simply $$sales_t = \alpha +\beta_1{advrtsng} +\epsilon_t$$. Contemporaneous advertising is highly significant.

```{r}
model1<-lm(sales~advrtsng);summary(model1)
```

Residual plot
```{r}
plot(ts(resid(model1),start=1907),xlab="year",ylab="residual",main="mo
del 1 residuals vs. year")
```

Adding a factor variable for advcopy.

```{r}
fadvcopy<-as.factor(advcopy)
model2<-lm(sales~advrtsng+fadvcopy);summary(model2)
```


```{r}
 model3<-lm(sales~advrtsng+Ind1+Ind2+Ind3);summary(model3)
```


```{r}
lag2adv<-c(NA,NA,head(advrtsng,-2))
lag3adv<-c(NA,NA,NA,head(advrtsng,-3))
lag4adv<-c(NA,NA,NA,NA,head(advrtsng,-4))
lag5adv<-c(NA,NA,NA,NA,NA,head(advrtsng,-5))
plot(lagadv,sales)
abline(lm(sales~lagadv))
plot(lag2adv,sales)
abline(lm(sales~lag2adv))
plot(lag3adv,sales)
abline(lm(sales~lag3adv))
plot(lag4adv,sales)
abline(lm(sales~lag4adv))
```

Fit a distributed lag model with geometric decay

$$
S_t = \alpha + \beta(A_t + \gamma{A_{t-1}} + \gamma^2{A_{t-2}} + \gamma^3{A_{t-3}}) + \epsilon_t \\
= \alpha +\beta(1-\gamma{\beta})^{-1}A_{t} + \epsilon_t
$$

where B is the backward shift operator. Recall that this formulation is equivalent to the autoregressive representation

$$
S_t = (1-\gamma)\alpha+\gamma{S_{t-1}} + \beta{A_t} + \epsilon_t - \gamma{\epsilon}_{t-1}
$$

In this model the contemporaenous effect of advertising is $\Beta$, the cumulative effect after one more time period is $\beta(1+\gamma)$, and the cumulatibe effect after m-1 more time periods is 

$$
\beta(1+\lambda + \lambda^2+...+\lambda^{m-1}) = \beta(1-\lambda^{m})/(1-\lambda)
$$

As m increases, this converged to $\beta/(1-\lambda)$. Thus, the proportion fo the long-run cumulative effect of advertising upon sales realized after m time periods is $1-\lambda^m$. Given p, the solution for m is

$$
m = log_e(1-p)/log_ey
$$

The model below is 

$$
S_t = \alpha+\beta_1{A_t} + \beta_2{S_{t-1}} + \beta_3{Ind1_t} + \beta_4{Ind2_t}
+ \beta_5{Ind3_t} + \epsilon_t
$$

```{r}
lagsales<-c(NA,head(sales,-1))
lag2sales<-c(NA,NA,head(sales,-2))
model4<-lm(sales~lagsales+advrtsng+Ind1+Ind2+Ind3);summary(model4)
```

Add lag two of sales as an explanatory variable

```{r}
model5<-
lm(sales~lagsales+lag2sales+advrtsng+Ind1+Ind2+Ind3);summary(model5)
```

Fit another model. Add lag one advertising to the precious fit

```{r}
options(digits=9)
model6<-
lm(sales~lagsales+lag2sales+advrtsng+lagadv+Ind1+Ind2+Ind3);summary(model6)
plot(ts(resid(model6),start=c(1907,1)),xlab="year",ylab="residual",main="model 6 residuals vs. year")
```


Calculate the 90% duration interval. Determine the first two deltas by hand calculation

```{r}
deltapartial<-delta<-c(rep(0,times=500))
#deltapartial is the partial sum of the deltas
delta[1]<-0.549964;delta[2]<-0.186242
deltapartial[1]<-delta[1]
deltapartial[2]<-deltapartial[1]+delta[2]
for(j in 3:500){
j1<-j-1;j2<-j-2
delta[j]<-0.965412*delta[j1]-0.205023*delta[j2]
deltapartial[j]<-deltapartial[j1]+delta[j]
}
deltapartial[500]*0.9
deltapartial[1:20]

```



















