--- 
title: "Time Series Forecasting"
author: "Katherine Wilson"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
---

# basics

## Decomposition Models 

**Additive Decomposition Model**

Where y is the response, T is the trend component (slowly varying), S is a seasonal component, and epsilon is a disturbance term:

$$
\begin{equation}
  y_t = T_t + S_t + \epsilon_t    
\end{equation}
$$

**Multiplicative Decomposition Model**


$$
\begin{equation}
  y_t = T_t S_t \epsilon_t    
\end{equation}
$$

**Additive Model**

Logging converts the multiplicative form to additive in thel ogs of components. 

$$
\begin{equation}
  y_t = logT_t + logS_t + log\epsilon_t     
\end{equation}
$$

Use multiplicative model in cases of increasing volaticity as response rises, rather than additive. If volatility does not change according to response level, both additive and multiplicative can be used. 

## Model assumptions

Mean of 0 and constant variance for errors
Errors are uncorrelated.
Errors are normally distributed

## Examples

### Example


```{r eval=FALSE, include=TRUE}
usbeer <- read.csv("/cloud/project/data/beernew.txt")
attach(usbeer)
head(usbeer)
class(usbeer)
usbeer.ts<-ts(beer,start=c(1987,1),freq=12)
plot(usbeer.ts,ylab="production",main="monthly US beer production") 
cycle(usbeer.ts) 
boxplot(usbeer.ts~cycle(usbeer.ts)) 
fmonth<-as.factor(month) 
boxplot(usbeer.ts~fmonth) 
beerexpansion<- c(beer[1:43],rep(NA,8),beer[52:171],rep(NA,8),beer[180:252],rep(NA,18),
beer[271:372]) 
beercontraction<-
c(rep(NA,43),beer[44:51],rep(NA,120),beer[172:179],rep(NA,73),beer[253:
270],rep(NA,102)) 
plot(ts(beerexpansion,start=c(1987,1),freq=12),ylim=c(13,22),ylab="production",main="monthly US beer production",col="green",lwd=2) 

lines(ts(beercontraction,start=c(1987,1),freq=12),col="red",lwd=2) 
legend("topleft",legend=c("expansion","contraction"),col=c("green","red
"),lty=1,cex=0.8)
```

### Example

```{r eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
ontgas <- read.csv("/cloud/project/data/Ontariogasdemand.txt")
attach(ontgas)
# head(ontgas)
ontariogas.ts <- ts(ontgas[,1], start = c(1960,1), freq = 12)
plot(ontariogas.ts, ylab = "gas demand", main = "Ontario gas demand,
1960-1975")

```


### Example
```{r eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
memp <- read.csv("/cloud/project/data/Memp1619.txt")
attach(memp)
# head(memp)


```


### Example

```{r eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
pass <- read.csv("/cloud/project/data/Intair.txt")
attach(pass)
# head(pass)
pass.ts <- ts(passengers, start = c(1949,1), freq = 12)
plot(pass.ts, ylab = "passengers (1000s)", main = "monthly international airline passengers, in 1000s")
lpass <- log(passengers)
class(lpass)
lpass.ts <- ts(lpass, start = c(1949,1), freq = 12)
class(lpass.ts)
plot(lpass.ts, ylab = "log(passengers(1000s))", main = "log monthly international airline passengers, in 1000s")

monthavg <- matrix(rep(0,12), ncol = 1)
for(i in 1:12){
  monthavg[i] <- mean(window(pass.ts, start = c(1949,i), freq = TRUE))
}
monthavg
options(digits=5)
monthavg
month <- rep(seq(1:12), times =12)
matrix(tapply(passengers, month, mean), ncol = 1)
```

### Example
```{r eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
trappings <- read.csv("/cloud/project/data/Lynx.txt")
attach(trappings)
head(trappings)
trappings[28:33,]
lynx.ts<- ts(trappings[,1],start = 1821)
plot(lynx.ts, ylab = "trappings", main = "annual lynx trappings")
lynx.ts <- ts(trappings[,1], start = 1821)
mink.ts <- ts(trappings[,4], start = 1821)
muskrat.ts <- ts(trappings[,5], start = 1821)
plot(cbind(lynx.ts, mink.ts, muskrat.ts))
```


### Example

```{r eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
garbage <- read.csv("/cloud/project/data/Garbage (1).txt")
attach(garbage)
head(garbage)
tonnage.ts<-ts(garbage[,4])
plot(tonnage.ts,ylab = "tonnage", main = "weekly tonnage recorded")
```

### Example

```{r eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
rain <- read.csv("/cloud/project/data/rainfall.txt")
attach(rain)
head(rain)
plot(ts(rainfall, start = c(1979,1), freq =12))
```

### Example

```{r eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
nyseadv <- read.csv("/cloud/project/data/nyseadv.txt")
attach(nyseadv)
head(nyseadv)
pctadvnc.ts <- ts(pctadvnc)
plot(pctadvnc.ts, ylab = "percent advances", main = "NYSE daily percent advances")
abline(h=mean(pctadvnc), lty = 2, col = "red")
```

### Example

```{r eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
tsa <- read.csv("/cloud/project/data/TSAnew2.txt")
attach(tsa)
head(tsa)
tsa19.ts <- ts(Travelers[1:366], start = c(1), freq = 7)
plot(tsa19.ts, ylab = "Passengers", xlab = "Week", main = "Passenger throughput, 3/1/2019-2/29/2020")
tsa192.ts <- ts(Travelers[1:366])
plot(tsa192.ts, ylab = "Passengers", xlab = "Time",
     main = "Passenger throughout, 3/1/2019-2/29/2020")
abline(v=c(32,62,93,123,154,185,215,246,276,307,338),
       lty=2, col = "red")
tsa20.ts <- ts(Travelers[367:731],start = c(1), freq = 7)
plot(tsa20.ts, ylab = "Passengers", xlab = "Week", main= "Passenger throughput, 3/1/1010-2/28/2021")
ltsa20.ts <- ts(log(Travelers)[367:731],start = c(1), freq = 7)
plot(ltsa20.ts, ylab = "Passengers", xlab = "Week", main = "log Passenger throughput, 3/1/2020-2/28/2021")
ltsa19.ts <- ts(log(Travelers)[1:366],start = c(1), freq = 7)
plot(ltsa19.ts, ylab = "log Passengers", xlab = "Week", main = "Passenger throughput", ylim = c(11,16), lty = 1, lwd = 2, col= "red")
lines(ltsa20.ts, lty=1, lwd = 2, col = "blue")
legend("topleft", legend = c("3/1/2019-2/29/2020","3/1/2020-2/28/2021"), col = c("red", "blue"), lty = 1, cex = 0.9)

tsa21.ts <- ts(Travelers[673:1037],start = c(1), freq =7)
plot(tsa21.ts, ylab = "Passengers", xlab = "Week", main = "Passenger throughput, 1/1/2021-12/31/2021")
weekdayav19 <- tapply(Travelers[1:366],Weekday[1:366],mean)
dmweekdayav19 <- weekdayav19-mean(weekdayav19)
cbind(weekdayav19,dmweekdayav19)
weekdayav21 <- tapply(Travelers[673:1037],Weekday[673:1037], mean)
dmweekdayav21 <- weekdayav21-mean(weekdayav21)
cbind(weekdayav21, dmweekdayav21)
```



